---
title: "Joined PCA plots of GEL and 1kg samples"
output: html_document
editor_options: 
  chunk_output_type: console
---

started: Alexey Larionov, 01May2019  
last updated: Alexey Larionov, 31May2019  

# summary  

- Read eigenvectors and eigenvalues  
- Read phenotype annotations for 1kg samples (for ethnic populations)  
- Read phenotype annotations for GEL samples (for case-control status)  
- Add phenotypes to eigenvectors (for colour coding on plots)  
- Plot eigenvalues  
- Plot eighenvectors for all samples and for White: British/Irish only  
- Make a list of 8,628 samples for White: British/Irish only  

- Compare plots with the published by 1kg consortsium in paper "An integrated map of structural variation in 2,504 human genomes"  
  (Ext Data Figure 5)  


```{r start_section}

Sys.time()
rm(list=ls())
graphics.off()

# Libraries installed in system folders
library(dplyr)
library(ggplot2)
library(plotly)

base_folder <- "/home/alarionov/re_gecip/inherited_cancer_predisposition/tischkowitz/users/alexey/fa_analysis"

options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

```

# Read PCA data

```{r read_pca_data}

# Compile file names
pca_data_folder=paste(base_folder,"data/s13_joined_gel_1kg_PCA/s04_calculate_eigenvectors/p03_thinn_and_pca",sep="/") 
eigenvalues_file=paste(pca_data_folder,"gel_1kg_10kv_100PCs.eigenval",sep="/")
eigenvectors_file=paste(pca_data_folder,"gel_1kg_10kv_100PCs.eigenvec",sep="/")

# Read data
eigenval.df <- read.table(eigenvalues_file)
eigenvec.df <- read.table(eigenvectors_file, header = T)

# Check eigenvalues
dim(eigenval.df)
eigenval.df[1:10,]

# Check and update eigenvectors
dim(eigenvec.df)
eigenvec.df[1:5,1:5]

eigenvec.df <- eigenvec.df %>% 
  select(-FID) %>% 
  rename(sample=IID)

eigenvec.df[1:5,1:5]

rm(pca_data_folder, eigenvalues_file, eigenvectors_file)

```

# Read 1kg phenotypes

```{r read_1kg_phenotypes}

data_file <- "/home/alarionov/public_data_resources/1000-genomes/20130502_GRCh37/integrated_call_samples_v3.20130502.ALL.panel"
phenotypes_1kg.df <- read.table(data_file,header = T)

dim(phenotypes_1kg.df)
phenotypes_1kg.df[1:5,]

# Check for duplicates
sum(duplicated(phenotypes_1kg.df$sample))

# Clean-up
rm(data_file)

```

# Read gel phenotypes

```{r read_gel_phenotypes}

# Load
load(paste(base_folder,"data/s12_explore_and_filter/s06_explore_and_update_phenotypes.RData",sep="/"))

# Check
dim(phenotypes.df)
sum(duplicated(phenotypes.df$Platekey))

# Rname
phenotypes_gel.df <- phenotypes.df

# Clean-up
rm(phenotypes.df)

```

# Combime phenotypes

Note increased figure heght to fit the vertical fprmat of figure   

```{r combine_phenotypes, fig.height=7}

# Compile joined data fgrame with phenotypes

phenotypes_1kg.df <- phenotypes_1kg.df %>% select(sample, group=super_pop)
phenotypes_1kg.df <- cbind(phenotypes_1kg.df, gel_ethnicity="1kg")

phenotypes_gel.df <- phenotypes_gel.df %>% select(sample=Platekey, group=Group, gel_ethnicity=Participant.Ethnic.Category)

phenotypes.df <- rbind(phenotypes_1kg.df, phenotypes_gel.df)

head(phenotypes.df)
tail(phenotypes.df)

# Explore group column
table(phenotypes.df$group)

# Make table of ethnicity in the joined dataset
sort(table(phenotypes.df$gel_ethnicity), decreasing = T)

# --- Make plot of ethnic groups in the GEL subset of data --- #

# Adjust default margins (to make space at the left of the plot for outside legend)
init_mar <- par("mar") # save initial grafics parameters
par(mar=c(20,4,4,4)) # modify margins

# Plot
barplot(sort(table(phenotypes.df[phenotypes.df$gel_ethnicity!="1kg","gel_ethnicity"]), 
             decreasing = T),
        las=2, main="Reported  ethnicity in selected GEL samples")

# Restore initial graphics parameters
par(mar=init_mar)

# Clean-up
rm(phenotypes_1kg.df, phenotypes_gel.df, init_mar)

```

# Add phenotypes to eigenvectors

```{r add_phenotypes_to_eigenvectors}

eigenphen.df <- inner_join(phenotypes.df, eigenvec.df, by="sample")
dim(eigenphen.df)

head(eigenphen.df[,1:5])
tail(eigenphen.df[,1:5])

rm(phenotypes.df, eigenvec.df)

```

# Plot eigenvalues

```{r plot_eigenvalues}

plot(eigenval.df$V1, 
     main="GEL + 1kg PCA: Variance captured by top 100 eigenvectors",
     ylab="Variance", xlab="Eigenvector", type="b")

rm(eigenval.df)

```

# Plot PC1 vs PC2

### A simple plot

```{r simple_plot}

# Explore groups
table(eigenphen.df$group)

# Prepare vector of colours
myColours <- eigenphen.df$group
"blue" -> myColours[myColours=="EUR"]
"brown" -> myColours[myColours=="AFR"]
"green" -> myColours[myColours=="AMR"]
"grey" -> myColours[myColours=="SAS"]
"magenta" -> myColours[myColours=="EAS"]
"red" -> myColours[myColours=="cases"]
"pink" -> myColours[myColours=="controls"]

# Prepare vector of point types (4 is for cross, 19 is for circle)
myPch <- eigenphen.df$group
4 -> myPch[myPch=="EUR"]
4 -> myPch[myPch=="AFR"]
4 -> myPch[myPch=="AMR"]
4 -> myPch[myPch=="SAS"]
4 -> myPch[myPch=="EAS"]
19 -> myPch[myPch=="cases"]
19 -> myPch[myPch=="controls"]
myPch <- as.integer(myPch)

# Adjust default margins (to make space at the left of the plot for outside legend)
init_mar <- par("mar") # save initial grafics parameters
init_xpd <- par("xpd") # save initial grafics parameters
par(mar=c(4,4,4,8), xpd=T) # modify margins

# Make the plot
plot(PC2~PC1, data=eigenphen.df, 
     col=myColours, pch=myPch, 
     main="GEL and 1kg PCA plot")

# Add legend
legend("topright",inset=c(-0.3,0),
       legend=c("EUR","AFR","AMR","SAS","EAS","GEL cases","GEL controls"),
       pch=c(4,4,4,4,4,19,19),
       col=c("blue","brown","green","grey","magenta","red","pink"),
       bty="n")

# Restore initial graphics parameters
par(mar=init_mar, xpd=init_xpd)

# Clean-up
rm(myColours, myPch, init_mar, init_xpd)

```

### ggplot

```{r ggplot}

# Prepare named vector for colour scale 
myColours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", 
               "SAS" = "GREY", "EAS" = "MAGENTA", 
               "cases" = "RED", "controls" = "PINK")
myColours

# Convert the colour vector to format required by ggplot
myColourScale <- scale_colour_manual(values=myColours) 

# Make plot
# note that dots are plotted in order of the data, and legend is in alphabetic order
ggplot(eigenphen.df, aes(PC1, PC2)) +
  geom_point(aes(col=group)) + 
  labs(title="GEL and 1kg PCA plot", x = "PC1", y ="PC2") +
  myColourScale

# Clean-up (keep myColourScale for later use)
rm(myColours)

```

### ggplotly

```{r ggplotly}

# Prepare vector with cases IDs
cases_ids <- eigenphen.df[,"sample"]

# Prepare factor with cases labels (it has to be a factor, not vector)
cases_labels <- as.factor(eigenphen.df$group)

# Explicitly reset order of levels in factor: this will define the order of legend and plotting (last elements are plotted on top)
cases_labels <- factor(cases_labels, levels=c("AMR", "SAS", "EAS", "EUR", "AFR", "cases", "controls"))

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen.df, aes(PC1, PC2)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL and 1kg PCA plot", x = "PC1", y ="PC2") +
  myColourScale

# Render plor with ggplotly
ggplotly(g)

# Clean-up
rm(g, myColourScale, cases_ids, cases_labels)

```

# ggplot PCA plot including the White British GEL participants only

```{r}

selected_samples <- eigenphen.df$gel_ethnicity == "1kg" | 
                    eigenphen.df$gel_ethnicity == "White: British" | 
                    eigenphen.df$gel_ethnicity == "White: Irish"
sum(selected_samples)

eigenphen.df <- eigenphen.df[selected_samples,]
dim(eigenphen.df)

# Prepare named vector for colour scale 
myColours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", 
               "SAS" = "GREY", "EAS" = "MAGENTA", 
               "cases" = "RED", "controls" = "PINK")
myColours

# Convert the colour vector to format required by ggplot
myColourScale <- scale_colour_manual(values=myColours) 

# Make plot
# note that dots are plotted in order of the data, and legend is in alphabetic order
ggplot(eigenphen.df, aes(PC1, PC2)) +
  geom_point(aes(col=group)) + 
  labs(title="GEL and 1kg PCA plot", x = "PC1", y ="PC2") +
  myColourScale

# Clean-up (keep myColourScale for later use)
rm(myColours, myColourScale, selected_samples)

```

# plotly PCA plot including the White British GEL participants only

```{r}

# Prepare named vector for colour scale 
myColours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", 
               "SAS" = "GREY", "EAS" = "MAGENTA", 
               "cases" = "RED", "controls" = "PINK")

# Convert the colour vector to format required by ggplot
myColourScale <- scale_colour_manual(values=myColours) 

# Prepare vector with cases IDs
cases_ids <- eigenphen.df[,"sample"]

# Prepare factor with cases labels (it has to be a factor, not vector)
cases_labels <- as.factor(eigenphen.df$group)

# Explicitly reset order of levels in factor: this will define the order of legend and plotting (last elements are plotted on top)
cases_labels <- factor(cases_labels, levels=c("AMR", "SAS", "EAS", "EUR", "AFR", "cases", "controls"))

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen.df, aes(PC1, PC2)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL and 1kg PCA plot", x = "PC1", y ="PC2") +
  myColourScale

# Render plor with ggplotly
ggplotly(g)

# Clean-up
rm(g, myColourScale, myColours, cases_ids, cases_labels)

```

# Make list of 8,628 samples for downstream analysis

```{r}

# Cjeck ethnic groups still retained in the current data
table(eigenphen.df$gel_ethnicity)

# Select cases (note drop=F: this is to keep output as a DF, rather than a vector)
selected_samples <- eigenphen.df[eigenphen.df$gel_ethnicity != "1kg", "sample", drop=F]
dim(selected_samples)

# Write cases into the text file
write.table(selected_samples, 
            file=paste(base_folder,"data/s13_joined_gel_1kg_PCA/s05_make_PCA_plots/reported_white_british_irish.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

# Clean-up
rm(selected_samples)

```

# save data

```{r}

save(eigenphen.df, 
     file=paste(base_folder,"data/s13_joined_gel_1kg_PCA/s05_make_PCA_plots/s01_plot_gel_1kg_PCA.RData",sep="/"))

```

# final section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
