---
title: "PCA plots of GEL only samples"
output: html_document
editor_options: 
  chunk_output_type: console
---

started: Alexey Larionov, 01May2019  
last updated: Carmelina Charalambous 04June2019  

# summary  

- Read eigenvectors and eigenvalues  
- Read phenotype annotations (for case-control status)  
- Add phenotypes to eigenvectors (for colour coding on plots)  
- Plot eigenvalues  
- Plot eighenvectors  
- Select outliers, print to the text file for later use  

```{r start_section}

Sys.time()
rm(list=ls())
graphics.off()

# Libraries installed in system folders
library(dplyr)
library(ggplot2)
library(plotly)

base_folder <- "/home/ccharalambous/re_gecip/inherited_cancer_predisposition/ccharalambous/Final_execution"

options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

```

# Read PCA data

```{r read_pca_data}

# Compile file names
pca_data_folder=paste(base_folder,"data/S14_gel_only_PCA/s02_calculate_eigenvectors/p03_thinn_and_pca",sep="/") 
eigenvalues_file=paste(pca_data_folder,"gel_only_10kv_100PCs.eigenval",sep="/")
eigenvectors_file=paste(pca_data_folder,"gel_only_10kv_100PCs.eigenvec",sep="/")

# Read data
eigenval.df <- read.table(eigenvalues_file)
eigenvec.df <- read.table(eigenvectors_file, header = T)

# Check eigenvalues
dim(eigenval.df)
eigenval.df[1:10,]

# Check and update eigenvectors
dim(eigenvec.df)
eigenvec.df[1:5,1:5]

eigenvec.df <- eigenvec.df %>% 
  select(-FID) %>% 
  rename(sample=IID)

eigenvec.df[1:5,1:5]

rm(pca_data_folder, eigenvalues_file, eigenvectors_file)

```

# Read phenotypes

```{r read_phenotypes}

# Load
load(paste(base_folder,"data/S12_explore_and_filter/s06_explore_and_update_phenotypes.RData",sep="/"))

# Check
dim(phenotypes.df)
sum(duplicated(phenotypes.df$Platekey))

```

# Explore ethnicities

```{r explore_ethnicities, fig.height=7}

# Explore group column
table(phenotypes.df$Group)

# Make table of ethnicity in the joined dataset
sort(table(phenotypes.df$Participant.Ethnic.Category), decreasing = T)

# --- Make plot of ethnic groups in the GEL subset of data --- #

# Adjust default margins (to make space at the left of the plot for outside legend)
init_mar <- par("mar") # save initial grafics parameters
par(mar=c(20,4,4,4)) # modify margins

# Plot
barplot(sort(table(phenotypes.df$Participant.Ethnic.Category), 
             decreasing = T),
        las=2, main="Reported  ethnicity in selected GEL samples")

# Restore initial graphics parameters
par(mar=init_mar)

# Clean-up
rm(init_mar)

```

# Add phenotypes to eigenvectors

```{r add_phenotypes_to_eigenvectors}

phenotypes.df <- phenotypes.df %>% rename(sample=Platekey)

eigenphen.df <- inner_join(phenotypes.df, eigenvec.df, by="sample")
dim(eigenphen.df)

head(eigenphen.df[,1:5])
tail(eigenphen.df[,1:5])

rm(phenotypes.df, eigenvec.df)

```

# Plot eigenvalues

```{r plot_eigenvalues}

plot(eigenval.df$V1, 
     main="GEL only PCA: Variance captured by top 100 eigenvectors",
     ylab="Variance", xlab="Eigenvector", type="b")

rm(eigenval.df)

```

# Plot PC1 to PC5

### A simple plot

```{r simple_plot}

# Explore groups
table(eigenphen.df$Group)

# Prepare vector of colours
myColours <- eigenphen.df$Group
"red" -> myColours[myColours=="cases"]
"blue" -> myColours[myColours=="controls"]

# Adjust default margins (to make space at the left of the plot for outside legend)
init_mar <- par("mar") # save initial grafics parameters
init_xpd <- par("xpd") # save initial grafics parameters
par(mar=c(4,4,4,8), xpd=T) # modify margins

# --- Make PC2 vs PC1 plot --- #

plot(PC2~PC1, data=eigenphen.df, 
     col=myColours, pch=4, 
     main="GEL only PCA plot")

# Add threshold lines at 3 x sd
lo_th_pc1 <- mean(eigenphen.df$PC1) - 3 * sd(eigenphen.df$PC1)
hi_th_pc1 <- mean(eigenphen.df$PC1) + 3 * sd(eigenphen.df$PC1)

lo_th_pc2 <- mean(eigenphen.df$PC2) - 3 * sd(eigenphen.df$PC2)
hi_th_pc2 <- mean(eigenphen.df$PC2) + 3 * sd(eigenphen.df$PC2)

abline(v=lo_th_pc1, lty=2, xpd=F)
abline(v=hi_th_pc1, lty=2, xpd=F)
abline(h=lo_th_pc2, lty=2, xpd=F)
abline(h=hi_th_pc2, lty=2, xpd=F)

# Add legend
legend("topright",inset=c(-0.2,0),
       legend=c("cases","controls"),
       pch=4,
       col=c("red","blue"),
       bty="n")

# --- Make PC3 vs PC2 plot --- #

plot(PC3~PC2, data=eigenphen.df, 
     col=myColours, pch=4, 
     main="GEL only PCA plot")

# Add threshold lines at 3 x sd
lo_th_pc3 <- mean(eigenphen.df$PC3) - 3 * sd(eigenphen.df$PC3)
hi_th_pc3 <- mean(eigenphen.df$PC3) + 3 * sd(eigenphen.df$PC3)

abline(v=lo_th_pc2, lty=2, xpd=F)
abline(v=hi_th_pc2, lty=2, xpd=F)
abline(h=lo_th_pc3, lty=2, xpd=F)
abline(h=hi_th_pc3, lty=2, xpd=F)

# Add legend
legend("topright",inset=c(-0.2,0),
       legend=c("cases","controls"),
       pch=4,
       col=c("red","blue"),
       bty="n")

# --- Make PC4 vs PC3 plot --- #

plot(PC4~PC3, data=eigenphen.df, 
     col=myColours, pch=4, 
     main="GEL only PCA plot")

# Add threshold lines at 3 x sd
lo_th_pc4 <- mean(eigenphen.df$PC4) - 3 * sd(eigenphen.df$PC4)
hi_th_pc4 <- mean(eigenphen.df$PC4) + 3 * sd(eigenphen.df$PC4)

abline(v=lo_th_pc3, lty=2, xpd=F)
abline(v=hi_th_pc3, lty=2, xpd=F)
abline(h=lo_th_pc4, lty=2, xpd=F)
abline(h=hi_th_pc4, lty=2, xpd=F)

# Add legend
legend("topright",inset=c(-0.2,0),
       legend=c("cases","controls"),
       pch=4,
       col=c("red","blue"),
       bty="n")

# --- Make PC5 vs PC4 plot --- #

plot(PC5~PC4, data=eigenphen.df, 
     col=myColours, pch=4, 
     main="GEL only PCA plot")

# Add threshold lines at 3 x sd
lo_th_pc5 <- mean(eigenphen.df$PC5) - 3 * sd(eigenphen.df$PC5)
hi_th_pc5 <- mean(eigenphen.df$PC5) + 3 * sd(eigenphen.df$PC5)

abline(v=lo_th_pc4, lty=2, xpd=F)
abline(v=hi_th_pc4, lty=2, xpd=F)
abline(h=lo_th_pc5, lty=2, xpd=F)
abline(h=hi_th_pc5, lty=2, xpd=F)

# Add legend
legend("topright",inset=c(-0.2,0),
       legend=c("cases","controls"),
       pch=4,
       col=c("red","blue"),
       bty="n")

# Restore initial graphics parameters
par(mar=init_mar,xpd=init_xpd)

# Clean-up
rm(myColours, init_mar, init_xpd,
   lo_th_pc1, lo_th_pc2, lo_th_pc3, lo_th_pc4, lo_th_pc5, 
   hi_th_pc1, hi_th_pc2, hi_th_pc3, hi_th_pc4, hi_th_pc5)

```

### ggplot

```{r ggplot}

# Prepare named vector for colour scale 
myColours <- c("cases" = "red", "controls" = "blue")
myColours

# Convert the colour vector to format required by ggplot
myColourScale <- scale_colour_manual(values=myColours) 

# Make plot
# note that dots are plotted in order of the data, and legend is in alphabetic order
ggplot(eigenphen.df, aes(PC1, PC2)) +
  geom_point(aes(col=Group)) + 
  labs(title="GEL only PCA plot", x = "PC1", y ="PC2") +
  myColourScale

# Clean-up
rm(myColours, myColourScale)

```

### ggplotly

```{r ggplotly}

# Prepare colour scale 
myColours <- c("cases" = "red", "controls" = "blue")
myColourScale <- scale_colour_manual(values=myColours) 

# Prepare vector with cases IDs
cases_ids <- eigenphen.df[,"sample"]

# Prepare factor with cases labels (it has to be a factor, not vector)
cases_labels <- as.factor(eigenphen.df$Group)

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen.df, aes(PC1, PC2)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL only PCA plot", x = "PC1", y ="PC2") +
  myColourScale

# Render plor with ggplotly
ggplotly(g)

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen.df, aes(PC2, PC3)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL only PCA plot", x = "PC2", y ="PC3") +
  myColourScale

# Render plor with ggplotly
ggplotly(g)

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen.df, aes(PC3, PC4)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL only PCA plot", x = "PC3", y ="PC4") +
  myColourScale

# Render plor with ggplotly
ggplotly(g)

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen.df, aes(PC4, PC5)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL only PCA plot", x = "PC4", y ="PC5") +
  myColourScale

# Render plor with ggplotly
ggplotly(g)

# Clean-up
rm(g, myColourScale, myColours, cases_ids, cases_labels)

```

# Function to get outliers on top 10 PCs

Takes one parameter: the SD threshold  
Assumes eigenphen.df in the environment  
Adds to the environment the set of logical vectors for outliers:  
lo_pc1, hi_pc1 ... lo_pc10, hi_pc10  

Note using <<- for making variables visible outside of the function  

```{r}

get_outliers.udf <- function(sd_threshold){
  
    # Start message
    cat("Selecting outliers in top 10 PCs at threshold",sd_threshold,"\n")
    
    # PC1 outliers
    pc1 <- eigenphen.df$PC1
    pc1_mean <- mean(pc1)
    pc1_sd  <- sd(pc1)
    lo_pc1 <<- pc1 < pc1_mean - sd_threshold * pc1_sd
    hi_pc1 <<- pc1 > pc1_mean + sd_threshold * pc1_sd
    cat("pc1 lo/hi:",sum(lo_pc1),"/",sum(hi_pc1),"\n")
    
    # PC2 outliers
    pc2 <- eigenphen.df$PC2
    pc2_mean <- mean(pc2)
    pc2_sd  <- sd(pc2)
    lo_pc2 <<- pc2 < pc2_mean - sd_threshold * pc2_sd
    hi_pc2 <<- pc2 > pc2_mean + sd_threshold * pc2_sd
    cat("pc2 lo/hi:",sum(lo_pc2),"/",sum(hi_pc2),"\n")
    
    # PC3 outliers
    pc3 <- eigenphen.df$PC3
    pc3_mean <- mean(pc3)
    pc3_sd  <- sd(pc3)
    lo_pc3 <<- pc3 < pc3_mean - sd_threshold * pc3_sd
    hi_pc3 <<- pc3 > pc3_mean + sd_threshold * pc3_sd
    cat("pc3 lo/hi:",sum(lo_pc3),"/",sum(hi_pc3),"\n")
    
    # PC4 outliers
    pc4 <- eigenphen.df$PC4
    pc4_mean <- mean(pc4)
    pc4_sd  <- sd(pc4)
    lo_pc4 <<- pc4 < pc4_mean - sd_threshold * pc4_sd
    hi_pc4 <<- pc4 > pc4_mean + sd_threshold * pc4_sd
    cat("pc4 lo/hi:",sum(lo_pc4),"/",sum(hi_pc4),"\n")
    
    # PC5 outliers
    pc5 <- eigenphen.df$PC5
    pc5_mean <- mean(pc5)
    pc5_sd  <- sd(pc5)
    lo_pc5 <<- pc5 < pc5_mean - 3 * pc5_sd
    hi_pc5 <<- pc5 > pc5_mean + 3 * pc5_sd
    cat("pc5 lo/hi:",sum(lo_pc5),"/",sum(hi_pc5),"\n")
    
    # PC6 outliers
    pc6 <- eigenphen.df$PC6
    pc6_mean <- mean(pc6)
    pc6_sd  <- sd(pc6)
    lo_pc6 <<- pc6 < pc6_mean - 6 * pc6_sd
    hi_pc6 <<- pc6 > pc6_mean + 6 * pc6_sd
    cat("pc6 lo/hi:",sum(lo_pc6),"/",sum(hi_pc6),"\n")
    
    # PC7 outliers
    pc7 <- eigenphen.df$PC7
    pc7_mean <- mean(pc7)
    pc7_sd  <- sd(pc7)
    lo_pc7 <<- pc7 < pc7_mean - 6 * pc7_sd
    hi_pc7 <<- pc7 > pc7_mean + 6 * pc7_sd
    cat("pc7 lo/hi:",sum(lo_pc7),"/",sum(hi_pc7),"\n")
    
    # PC8 outliers
    pc8 <- eigenphen.df$PC8
    pc8_mean <- mean(pc8)
    pc8_sd  <- sd(pc8)
    lo_pc8 <<- pc8 < pc8_mean - 6 * pc8_sd
    hi_pc8 <<- pc8 > pc8_mean + 6 * pc8_sd
    cat("pc8 lo/hi:",sum(lo_pc8),"/",sum(hi_pc8),"\n")
    
    # PC9 outliers
    pc9 <- eigenphen.df$PC9
    pc9_mean <- mean(pc9)
    pc9_sd  <- sd(pc9)
    lo_pc9 <<- pc9 < pc9_mean - 6 * pc9_sd
    hi_pc9 <<- pc9 > pc9_mean + 6 * pc9_sd
    cat("pc9 lo/hi:",sum(lo_pc9),"/",sum(hi_pc9),"\n")
    
    # PC10 outliers
    pc10 <- eigenphen.df$PC10
    pc10_mean <- mean(pc10)
    pc10_sd  <- sd(pc10)
    lo_pc10 <<- pc10 < pc10_mean - 6 * pc10_sd
    hi_pc10 <<- pc10 > pc10_mean + 6 * pc10_sd
    cat("pc10 lo/hi:",sum(lo_pc10),"/",sum(hi_pc10),"\n\n")
    
}

```

# Select eigenvectors outliers at 6 x sd

```{r}

# Get outliers at 6 x sd
get_outliers.udf(6)

# Combine outliers

ev_outliers_6sd_3pc <- lo_pc1 | lo_pc2 | lo_pc3 | 
                       hi_pc1 | hi_pc2 | hi_pc3

ev_outliers_6sd_5pc <- lo_pc1 | lo_pc2 | lo_pc3 | lo_pc4 | lo_pc5 | 
                       hi_pc1 | hi_pc2 | hi_pc3 | hi_pc4 | hi_pc5

ev_outliers_6sd_10pc <- lo_pc1 | lo_pc2 | lo_pc3 | lo_pc4 | lo_pc5 | lo_pc6 | lo_pc7 | lo_pc8 | lo_pc9 | lo_pc10 | 
                        hi_pc1 | hi_pc2 | hi_pc3 | hi_pc4 | hi_pc5 | hi_pc6 | hi_pc7 | hi_pc8 | hi_pc9 | hi_pc10

sum(ev_outliers_6sd_3pc) # 31 outliers @ 6 x sd on 3 top PCs
sum(ev_outliers_6sd_5pc) # 109 outliers @ 6 x sd on 5 top PCs
sum(ev_outliers_6sd_10pc) # 122 outliers @ 6 x sd on 10 top PCs

# Check the count and ethnicity of retained samples (w/o outliers @ 6 x sd on 5 top PCs)
sum(! ev_outliers_6sd_10pc) # 8,506 ethnically similar samples retained 
table(eigenphen.df[,"Participant.Ethnic.Category"]) # 8,498 British and 130 Irish
table(eigenphen.df[! ev_outliers_6sd_10pc,"Participant.Ethnic.Category"]) # 8,379 British and 127 Irish

# Get the samples names
ev_outliers_6sd_3pc.df <- eigenphen.df[ev_outliers_6sd_3pc, "sample", drop=F]
ev_outliers_6sd_5pc.df <- eigenphen.df[ev_outliers_6sd_5pc, "sample", drop=F]
ev_outliers_6sd_10pc.df <- eigenphen.df[ev_outliers_6sd_10pc, "sample", drop=F]

# Write selected samples into the text file
write.table(ev_outliers_6sd_3pc.df, 
            file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/ev_outliers_6sd_3pc.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

write.table(ev_outliers_6sd_5pc.df, 
            file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/ev_outliers_6sd_5pc.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

write.table(ev_outliers_6sd_10pc.df, 
            file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/ev_outliers_6sd_10pc.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

# Clean-up
rm(lo_pc1,lo_pc2,lo_pc3,lo_pc4,lo_pc5,lo_pc6,lo_pc7,lo_pc8,lo_pc9,lo_pc10,
   hi_pc1,hi_pc2,hi_pc3,hi_pc4,hi_pc5,hi_pc6,hi_pc7,hi_pc8,hi_pc9,hi_pc10,
   ev_outliers_6sd_3pc,ev_outliers_6sd_5pc,ev_outliers_6sd_10pc,
   ev_outliers_6sd_3pc.df,ev_outliers_6sd_5pc.df,ev_outliers_6sd_10pc.df)

```

# Select eigenvectors outliers at 3 x sd

```{r}

# Get outliers at 3 x sd
get_outliers.udf(3)

# Combine outliers

ev_outliers_3sd_3pc <- lo_pc1 | lo_pc2 | lo_pc3 | 
                       hi_pc1 | hi_pc2 | hi_pc3

ev_outliers_3sd_5pc <- lo_pc1 | lo_pc2 | lo_pc3 | lo_pc4 | lo_pc5 | 
                       hi_pc1 | hi_pc2 | hi_pc3 | hi_pc4 | hi_pc5

ev_outliers_3sd_10pc <- lo_pc1 | lo_pc2 | lo_pc3 | lo_pc4 | lo_pc5 | lo_pc6 | lo_pc7 | lo_pc8 | lo_pc9 | lo_pc10 | 
                        hi_pc1 | hi_pc2 | hi_pc3 | hi_pc4 | hi_pc5 | hi_pc6 | hi_pc7 | hi_pc8 | hi_pc9 | hi_pc10

sum(ev_outliers_3sd_3pc) # 144 outliers @ 3 x sd on 3 top PCs
sum(ev_outliers_3sd_5pc) # 236 outliers @ 3 x sd on 5 top PCs
sum(ev_outliers_3sd_10pc) # 249 outliers @ 3 x sd on 10 top PCs

# Check the count and ethnicity of retained samples (w/o outliers @ 3 x sd on 5 top PCs)
sum(! ev_outliers_3sd_5pc) # 8,392 ethnically similar samples retained 
table(eigenphen.df[,"Participant.Ethnic.Category"]) # 8,498 British and 130 Irish
table(eigenphen.df[! ev_outliers_3sd_5pc,"Participant.Ethnic.Category"]) # 8,269 British and 123 Irish

# Get the samples names
ev_outliers_3sd_3pc.df <- eigenphen.df[ev_outliers_3sd_3pc, "sample", drop=F]
ev_outliers_3sd_5pc.df <- eigenphen.df[ev_outliers_3sd_5pc, "sample", drop=F]
ev_outliers_3sd_10pc.df <- eigenphen.df[ev_outliers_3sd_10pc, "sample", drop=F]

# Just in case - the selected samples (based on later analysis)
selected_samples_3sd_5pc.df<- eigenphen.df[!ev_outliers_3sd_5pc, "sample", drop=F]

# Write selected samples into the text file
write.table(ev_outliers_3sd_3pc.df, 
            file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/ev_outliers_3sd_3pc.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

write.table(ev_outliers_3sd_5pc.df, 
            file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/ev_outliers_3sd_5pc.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

write.table(ev_outliers_3sd_10pc.df, 
            file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/ev_outliers_3sd_10pc.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

# 3sd_5pc was selected as the threshold after estimation of the PCA plots in the next step s14-s02
write.table(selected_samples_3sd_5pc.df, 
            file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/selected_samples_3sd_5pc.txt",sep="/"),
            col.names=F, row.names=F, quote=F)

# Clean-up
rm(lo_pc1,lo_pc2,lo_pc3,lo_pc4,lo_pc5,lo_pc6,lo_pc7,lo_pc8,lo_pc9,lo_pc10,
   hi_pc1,hi_pc2,hi_pc3,hi_pc4,hi_pc5,hi_pc6,hi_pc7,hi_pc8,hi_pc9,hi_pc10,
   ev_outliers_3sd_3pc,ev_outliers_3sd_5pc,ev_outliers_3sd_10pc,
   ev_outliers_3sd_3pc.df,ev_outliers_3sd_5pc.df,ev_outliers_3sd_10pc.df,
   selected_samples_3sd_5pc.df,get_outliers.udf)

```

# save data

```{r}

save(eigenphen.df, 
     file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/s01_plot_gel_only_PCA.RData",sep="/"))

```

# final section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
