---
title: "Plot selected samples in joined gel 1kg PCA"
output: html_document
editor_options: 
  chunk_output_type: console
---

started: Alexey Larionov, 29May2019  
last updated: Carmelina Charalambous 04June2019  

# summary  

- Read joined gel 1kg eigenvectors  
- Read ev outliers for 3sd: 3pc, 5pc and 10 pc  
- Plot PC1 vs PC2 w/o the outliers  
- 3sd 5pc has been selected as the threshold to remove ethnic outliers  

6sd was nit ploteed because it is clear from the counts of outliers that 6sd would be too relaxed  

```{r start_section}

Sys.time()
rm(list=ls())
graphics.off()

# Libraries installed in system folders
library(dplyr)
library(ggplot2)
library(plotly)

base_folder <- "/home/ccharalambous/re_gecip/inherited_cancer_predisposition/ccharalambous/Final_execution"

options(stringsAsFactors = F,
        warnPartialMatchArgs = T, 
        warnPartialMatchAttr = T, 
        warnPartialMatchDollar = T)

```

# Read joined PCA data

```{r read_pca_data}

load(paste(base_folder,"data/S13_add_eigenvectors/s05_make_PCA_plots/s01_plot_gel_1kg_PCA.RData",sep="/"))
dim(eigenphen.df)

```

# Read gel ev outliers

```{r read_phenotypes}

# Compile file name
data_folder <- paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots",sep="/") 

# Read outliers
ev_outliers_3sd_3pc.df <- read.table(paste(data_folder, "ev_outliers_3sd_3pc.txt", sep="/")
                                     , sep="\t", header=F, quote="")

ev_outliers_3sd_5pc.df <- read.table(paste(data_folder, "ev_outliers_3sd_5pc.txt", sep="/")
                                     , sep="\t", header=F, quote="")

ev_outliers_3sd_10pc.df <- read.table(paste(data_folder, "ev_outliers_3sd_10pc.txt", sep="/")
                                     , sep="\t", header=F, quote="")

dim(ev_outliers_3sd_3pc.df)
dim(ev_outliers_3sd_5pc.df)
dim(ev_outliers_3sd_10pc.df)

# Clean-up
rm(data_folder)

```

# Exclude outliers

```{r}

# Eexclude
eigenphen_3sd_3pc.df <- eigenphen.df[!eigenphen.df$sample %in% ev_outliers_3sd_3pc.df$V1,]
eigenphen_3sd_5pc.df <- eigenphen.df[!eigenphen.df$sample %in% ev_outliers_3sd_5pc.df$V1,]
eigenphen_3sd_10pc.df <- eigenphen.df[!eigenphen.df$sample %in% ev_outliers_3sd_10pc.df$V1,]

# Check result
table(eigenphen_3sd_3pc.df$gel_ethnicity)
table(eigenphen_3sd_5pc.df$gel_ethnicity)
table(eigenphen_3sd_10pc.df$gel_ethnicity)

# Clean-up
rm(ev_outliers_3sd_3pc.df, ev_outliers_3sd_5pc.df, ev_outliers_3sd_10pc.df, eigenphen.df)

```

# Plot PC1 vs PC2 for 3sd 3pc

```{r}

# Prepare named vector for colour scale 
myColours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", 
               "SAS" = "GREY", "EAS" = "MAGENTA", 
               "cases" = "RED", "controls" = "PINK")

# Convert the colour vector to format required by ggplot
myColourScale <- scale_colour_manual(values=myColours) 

# Prepare vector with cases IDs
cases_ids <- eigenphen_3sd_3pc.df[,"sample"]

# Prepare factor with cases labels (it has to be a factor, not vector)
cases_labels <- as.factor(eigenphen_3sd_3pc.df$group)

# Explicitly reset order of levels in factor: this will define the order of legend and plotting (last elements are plotted on top)
cases_labels <- factor(cases_labels, levels=c("AMR", "SAS", "EAS", "EUR", "AFR", "cases", "controls"))

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen_3sd_3pc.df, aes(PC1, PC2)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL and 1kg PCA plot: w/o 3sd on 3pc", x = "PC1", y ="PC2") +
  myColourScale

# Plot
g

# Render plot with ggplotly
ggplotly(g)

# Clean-up
rm(g, myColourScale, myColours, cases_ids, cases_labels)

```

# Plot PC1 vs PC2 for 3sd 5pc

```{r}

# Prepare named vector for colour scale 
myColours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", 
               "SAS" = "GREY", "EAS" = "MAGENTA", 
               "cases" = "RED", "controls" = "PINK")

# Convert the colour vector to format required by ggplot
myColourScale <- scale_colour_manual(values=myColours) 

# Prepare vector with cases IDs
cases_ids <- eigenphen_3sd_5pc.df[,"sample"]

# Prepare factor with cases labels (it has to be a factor, not vector)
cases_labels <- as.factor(eigenphen_3sd_5pc.df$group)

# Explicitly reset order of levels in factor: this will define the order of legend and plotting (last elements are plotted on top)
cases_labels <- factor(cases_labels, levels=c("AMR", "SAS", "EAS", "EUR", "AFR", "cases", "controls"))

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen_3sd_5pc.df, aes(PC1, PC2)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL and 1kg PCA plot: w/o 3sd on 5pc", x = "PC1", y ="PC2") +
  myColourScale

# Plot
g

# Render plot with ggplotly
ggplotly(g)

# Clean-up
rm(g, myColourScale, myColours, cases_ids, cases_labels)

```

# Plot PC1 vs PC2 for 3sd 10pc

```{r}

# Prepare named vector for colour scale 
myColours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", 
               "SAS" = "GREY", "EAS" = "MAGENTA", 
               "cases" = "RED", "controls" = "PINK")

# Convert the colour vector to format required by ggplot
myColourScale <- scale_colour_manual(values=myColours) 

# Prepare vector with cases IDs
cases_ids <- eigenphen_3sd_10pc.df[,"sample"]

# Prepare factor with cases labels (it has to be a factor, not vector)
cases_labels <- as.factor(eigenphen_3sd_10pc.df$group)

# Explicitly reset order of levels in factor: this will define the order of legend and plotting (last elements are plotted on top)
cases_labels <- factor(cases_labels, levels=c("AMR", "SAS", "EAS", "EUR", "AFR", "cases", "controls"))

# Make plot (save instead of immediate rendering)
g <- ggplot(eigenphen_3sd_10pc.df, aes(PC1, PC2)) +
  geom_point(aes(col=cases_labels, text=cases_ids)) + 
  labs(title="GEL and 1kg PCA plot: w/o 3sd on 10pc", x = "PC1", y ="PC2") +
  myColourScale

# Plot
g

# Render plot with ggplotly
ggplotly(g)

# Clean-up
rm(g, myColourScale, myColours, cases_ids, cases_labels)

```

# save data

```{r}

save(eigenphen_3sd_3pc.df,eigenphen_3sd_5pc.df, 
     file=paste(base_folder,"data/S14_gel_only_PCA/s03_make_PCA_plots/s02_plot_selected_samples_in_joined_gel_1kg_PCA.RData",sep="/"))

```

# final section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
